---
title: "Language Typology & Geography | Data Description"
author: "Language Typology Project Group"
date: "11/7/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load libraries
library(tidyverse)
library(maps)
```

```{r data load , echo=FALSE}
#import cleaned dataframe
load(file="../clean-data/by_language.RData")
```

# PART 1 | Describing Language Data

### TODO Language | Glottalized Consonants (Maho)

### Language | Consonant-Vowel Ratio (Maho)


**TODO: Explanation of what CV ratio is and why it's important**
Here is a visualization of languages with extreme high and low Consonant - Vowel ratios. 

```{r, results="hide"}

cvratio.low = by_language %>% filter(cv_ratio == '1 - Low')
cvratio.high = by_language %>% filter(cv_ratio == '5 - High')

#generate the map
world = map_data("world")
worldmap.combined.cvratio = ggplot(world, aes(x=long, y=lat, group=group))+
  geom_path()+
  scale_y_continuous(breaks=(-2:2) * 30)+
  scale_x_continuous(breaks=(-4:4) * 45)+
  geom_point(data = cvratio.low,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'blue',
             alpha = 0.8,
             size = 0.75)+
  geom_point(data = cvratio.high,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'red',
             alpha = 0.8,
             size = 0.9)+
  labs(title = "Map of Language with High(red) and Low(blue) Consonant-Value Ratios")
  theme_bw()
  
```

```{r}
#display the map
worldmap.combined.cvratio 
```



### TODO: Language | Tone (Ebru)

### TODO: Language | Speakers by Language (Amy)

Data for number of speakers by Lanaguage is remarkably difficult to locate! The primary aggregator source for this type of data is the Ethnologue service (https://www.ethnologue.com/), however, a costly subscription is required. Fortunately, we were able to locate a relevant Linguistics paper that used this dataset, and included their (cleaned) data as supplementary materials. For this, we offer thanks to Alexander Koplenig and The Royal Society Publishing Service. 

https://royalsocietypublishing.org/doi/10.1098/rsos.181274#d3e487				
https://royalsocietypublishing.org/doi/suppl/10.1098/rsos.181274


#### Distribution of Languages by Speaker Population
There are `r dim[by_language]` languages in our dataset, however, the number of humans that speak and understand these languages are widely different. In this figure, we see the distribution of languages by number of speakers as a histogram. The number of speakers are plotted on the x-axis, and number of languages within the bin-range on the y-axis. The top facet shows a linear scale, where we can see that a very large number of languages have very few speakers. The bottom facet plots the same data with a log10 scale, which is a better model for the speaker data. 
```{r speakers distribution}

options(scipen=10000) # print nums on scales not scientific notation
speakers <- by_language %>% select(wals_code, ISO639.3, name, population, L2prop) #get speaker data

plin <-  ggplot(data = speakers, aes(x=population)) +
  geom_histogram(alpha=0.5, bins=30)+
  labs(title="Distribution of Languages (linear scale)", 
       x = "Speaker Population", y="Number of Languages") +
  theme_minimal()

plog <- ggplot(data = speakers, aes(x=population)) +
  geom_histogram(alpha=0.5, bins=30)+
  labs(title="Distribution of Languages (log10 scale)", 
       x = "Speaker Population", y="Number of Languages") +
  scale_x_log10()+
  theme_minimal()

grid.arrange(plin, plog, bottom ="Distribution of Languages by Speaker Population")

remove(plin,plog)
```


```{r speakers distribution}

options(scipen=10000)

speakers <- by_language %>% select(wals_code, ISO639.3, name, population, L2prop) %>% 
arrange (-population) %>% mutate(
  second = population * L2prop,
  first = population - second
) %>% slice(1:25) 



plin <-  ggplot(data = speakers, aes(x=name, y=population)) +
  geom_bar(stat="identity")+
  labs(title="Number of Speakers for Top 20 Languages",
       x = "Language", y="Number of Speakers") +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 45))

plin
# 
# plog <- ggplot(data = speakers, aes(x=population)) +
#   geom_histogram(alpha=0.5, bins=30)+
#   labs(title="Distribution of Languages (log10 scale)", 
#        x = "Speaker Population", y="Number of Languages") +
#   scale_x_log10()+
#   theme_minimal()
# 
# grid.arrange(plin, plog, bottom ="Distribution of Languages by Speaker Population")


```
```


### TODO: Language | Elevation (Ebru)

### TODO: Language | Language Types (Cat)



# PART 2 | Describing Country Data

### TODO: Countries | GDP 
### TODO Countries | Linguistic Diversity 


# PART 3 | Describing Environmental Data

### TODO : temperature
### TODO : humidity