---
title: "PSYC 201A Project: Data Cleaning (Ejectives)"
author: "Maho Takahashi"
date: "2020-10-30"
output: html_document
---

## About this dataset

This dataset was retrived from the WALS database of glottalized consonants, including ejectives, of 567 languages (https://wals.info/feature/7A). In cleaning and visualizing the data, I used the following packages:    

```{r}
library(tidyverse)
library(maps)
```

Information in this dataset pertaining to our analysis is the following:

* `$wals.code`: 3-digit code unique to a language (e.g., English = eng) 

* `$description` and `$value`: Description of whether and what kind of glottalized consonants a language has, and its numerical coding
  + "No glottalized consonants" (value = 1): 409 rows
  + "Ejectives only" (value = 2): 58 rows
  + "Implosives only" (value = 3): 55 rows
  + "Glottalized resonants only" (value = 4): 4 rows
  + "Ejectives and implosives" (value = 5): 14 rows
  + "Ejectives and glottalized resonants" (value = 6): 20 rows
  + "Implosives and glottalized resonants" (value = 7): 4 rows
  + "Ejectives, implosives, and glottalized resonants" (value = 7): 3 rows

* `$latitude` and `$longitude`: Geographical information of a language

## Cleaning data
No cleaning was required for this dataset.

## Visualizing data

I'll first focus on languages without any glottalized consonants.

```{r}
ejectives.data = read.csv('ejectives.csv')
no.ejectives = ejectives.data %>% filter(description == 'No glottalized consonants')
```

The following code draws a world map using `geom_path`, and visualizes where languages without any glottalized consonants are located.

```{r}
world = map_data("world")
worldmap.with.no.glottalized.c = ggplot(world, aes(x=long, y=lat, group=group))+
  geom_path()+
  scale_y_continuous(breaks=(-2:2) * 30)+
  scale_x_continuous(breaks=(-4:4) * 45)+
  geom_point(data = no.ejectives,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'blue',
             size = 0.9)+
  theme_bw()
worldmap.with.no.glottalized.c
```

Next, I'll illustrate the locations of the languages with glottalized consonants.

```{r}
ejectives = ejectives.data %>% filter(!(description == 'No glottalized consonants'))
```

Here is the plot:

```{r}
worldmap.with.glottalized.c = ggplot(world, aes(x=long, y=lat, group=group))+
  geom_path()+
  scale_y_continuous(breaks=(-2:2) * 30)+
  scale_x_continuous(breaks=(-4:4) * 45)+
  geom_point(data = ejectives,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'red',
             size = 0.9)+
  theme_bw()
worldmap.with.glottalized.c
```

And here is the above two plots combined.

```{r}
worldmap.combined.ejectives = ggplot(world, aes(x=long, y=lat, group=group))+
  geom_path()+
  scale_y_continuous(breaks=(-2:2) * 30)+
  scale_x_continuous(breaks=(-4:4) * 45)+
  geom_point(data = no.ejectives,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'blue',
             alpha = 0.8,
             size = 0.75)+
  geom_point(data = ejectives,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'red',
             alpha = 0.8,
             size = 0.75)+
  theme_bw()
worldmap.combined.ejectives 
```


