---
title: "Complete Merge"
author: "Language Typology Group"
date: "11/15/2020"
output: html_document
---

##Move
Here, I added a column to connect the ISO alpha 2 country codes with the written-out country names.
```{r}
glimpse(by_language)
glimpse(by_country)
#country_code_iso <- read_csv("C:/Users/cathe/Downloads/country_code_iso.csv")
by_language <- by_language %>%
  separate(countrycodes, c("primary_country", "secondary_country"), "\\s")

country_language <- merge(by_language, country_code_iso, by.x = "primary_country", by.y = "Code", all = FALSE)
```

Here I will merge the GDP data with the new intermediary data frame `country_language`.

**It seems we lost 20 or so rows in the transition somehow - due to spelling differences?**

Manually add ISOcodes because of country name spelling differences

```{r}
str(country_language)
str(by_country)  #add 2-letter country code (ISO) "NEW_ISO"

unique(country_language$Name)
unique(by_country$Country)

by_country_language <- merge(country_language, by_country, by.x = "primary_country", by.y = "NEW_ISO")

colnames(by_country_language) <- c("country_name", "country_code", "wals_code", "X1", "iso_lang_code", "language_name", "genus", "subfamily", "family", "lsbranch", "ssbranch", "sbranch", "mbranch", "stock", "elev_ft", "elev_m", "latitude", "longitude", "secondary_country", "area", "macroarea", "continent", "glot_features", "cv_ratio", "tones", "rangesize", "population", "vehicularity", "L2prop", "H", "H_unigram", "lang.count", "lang.percent", "established.langs", "immigrant.langs", "total.speakers", "mean.speakers", "median.speakers", "diversity.index", "div.coverage", "GDP")

#write.csv(by_country_language, "by_country_language.csv")
save(by_country_language, file = "../clean-data/by_country_language.RData")