---
title: "PSYC 201A Project: Data Cleaning (C-V ratio)"
author: "Maho Takahashi"
date: "2020-10-30"
output: html_document
---

## About this dataset

This dataset was retrived from the WALS database of the consonant-vowel ratio of 564 languages (https://wals.info/feature/3A#). In cleaning and visualizing the data, I used the following packages:    

```{r}
library(tidyverse)
library(maps)
```

Information in this dataset relevant to our analysis is the following:

* `$wals.code`: 3-digit code unique to a language (e.g., English = eng) 

* `$description`: Description of the ratio between consonants and vowels, from "Low" (small # of consonants compared with vowels) to "High" (large # of consonants compared with vowels)  

* `$value`: Numerical coding of `$description`; 1 corresponds to "Low" and 5 corresponds to "High" 

* `$latitude` and `$longitude`: Geographical information of a language 

## Visualizing data

Here is a visualization of langauges with a "Low" C-V ratio.

```{r}
cvratio.data = read.csv('cvratio.csv')
cvratio.low = cvratio.data %>% filter(description == 'Low')
```

The following code draws a world map using `geom_path`, and visualizes where languages with a low C-V ratio are located.

```{r}
world = map_data("world")
worldmap.with.low.cv = ggplot(world, aes(x=long, y=lat, group=group))+
  geom_path()+
  scale_y_continuous(breaks=(-2:2) * 30)+
  scale_x_continuous(breaks=(-4:4) * 45)+
  geom_point(data = cvratio.low,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'blue',
             size = 0.9)+
  theme_bw()
worldmap.with.low.cv
```

Next, I'll illustrate the locations of the languages with a high C-V ratio.

```{r}
cvratio.high = cvratio.data %>% filter(description == 'High')
```

Here is the plot:

```{r}
world = map_data("world")
worldmap.with.high.cv = ggplot(world, aes(x=long, y=lat, group=group))+
  geom_path()+
  scale_y_continuous(breaks=(-2:2) * 30)+
  scale_x_continuous(breaks=(-4:4) * 45)+
  geom_point(data = cvratio.high,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'red',
             size = 0.9)+
  theme_bw()
worldmap.with.high.cv
```

And here is the above two plots combined.

```{r}
worldmap.combined.cvratio = ggplot(world, aes(x=long, y=lat, group=group))+
  geom_path()+
  scale_y_continuous(breaks=(-2:2) * 30)+
  scale_x_continuous(breaks=(-4:4) * 45)+
  geom_point(data = cvratio.low,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'blue',
             alpha = 0.8,
             size = 0.75)+
  geom_point(data = cvratio.high,
             aes(x=longitude, y=latitude),
             inherit.aes = FALSE,
             color = 'red',
             alpha = 0.8,
             size = 0.9)+
  theme_bw()
worldmap.combined.cvratio 
```