---
title: "Dataframe Merging"
author: "Language Typology Project Group"
date: "11/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#packages
library(tidyverse)
library(maps)
library(tibble)
```

# Tidying Dataframes and Metadata

## Ejectives

```{r, include=FALSE}
suppressMessages(ejectives <- read_csv("G:/Shared drives/PSYCH 201a project/Data/ejectives.csv"))
colnames(ejectives) <- c("wals_code", "name", "value", "description", "latitude", "longitude", "genus", "family", "area")
ejectives <- as.data.frame(ejectives)
write.csv(ejectives, "ejectives_cleaned.csv")
```

### About this dataset

This dataset was retrived from the WALS database of glottalized consonants, including ejectives, of 567 languages (https://wals.info/feature/7A). In cleaning and visualizing the data, I used the following packages:    

Information in this dataset pertaining to our analysis is the following:

* `$wals.code`: 3-digit code unique to a language (e.g., English = eng) 

* `$description` and `$value`: Description of whether and what kind of glottalized consonants a language has, and its numerical coding
  + "No glottalized consonants" (value = 1): 409 rows
  + "Ejectives only" (value = 2): 58 rows
  + "Implosives only" (value = 3): 55 rows
  + "Glottalized resonants only" (value = 4): 4 rows
  + "Ejectives and implosives" (value = 5): 14 rows
  + "Ejectives and glottalized resonants" (value = 6): 20 rows
  + "Implosives and glottalized resonants" (value = 7): 4 rows
  + "Ejectives, implosives, and glottalized resonants" (value = 7): 3 rows

* `$latitude` and `$longitude`: Geographical information of a language

```{r}
suppressMessages(ejectives_cleaned <- read_csv("G:/Shared drives/PSYCH 201a project/Data/ejectives_cleaned.csv"))
ejectives_cleaned <- ejectives_cleaned %>% select(-c(X1))
ejectives_cleaned <- as_data_frame(ejectives_cleaned)
#str(ejectives_cleaned)
```

## Consonant-Vowel Ratio

```{r, include=FALSE}
suppressMessages(CVratio <- read_csv("G:/Shared drives/PSYCH 201a project/Data/CVratio.csv"))
colnames(CVratio) <- c("wals_code", "name", "value", "description", "latitude", "longitude", "genus", "family", "area")
CVratio <- as.data.frame(CVratio)
write.csv(CVratio, "CVratio_cleaned.csv")
```

### About this dataset

This dataset was retrived from the WALS database of the consonant-vowel ratio of 564 languages (https://wals.info/feature/3A#).     

Information in this dataset relevant to our analysis is the following:

* `$wals.code`: 3-digit code unique to a language (e.g., English = eng) 

* `$description`: Description of the ratio between consonants and vowels, from "Low" (small # of consonants compared with vowels) to "High" (large # of consonants compared with vowels)  

* `$value`: Numerical coding of `$description`; 1 corresponds to "Low" and 5 corresponds to "High" 

* `$latitude` and `$longitude`: Geographical information of a language 

```{r}
suppressMessages(CVratio_cleaned <- read_csv("G:/Shared drives/PSYCH 201a project/Data/CVratio_cleaned.csv"))
CVratio_cleaned <- as_data_frame(CVratio_cleaned)
CVratio_cleaned <- CVratio_cleaned %>% select(-c(X1))
colnames(CVratio_cleaned) <- c("wals_code", "name", "value", "description", "latitude", "longitude", "genus", "family", "area")
str(CVratio_cleaned)
```

## Tone


### About this dataset

This dataset was retrived from the WALS database of the Tone of 527 languages (https://wals.info/feature/13A#).

* `$wals.code`: 3-digit code unique to a language (e.g., English = eng) 

* `$description`: The languages with tones are divided into those with a simple tone system — essentially those with only a two-way basic contrast, usually between high and low levels — and those with a more complex set of contrasts.

* `$value`: Numerical coding of `$description`; 1 corresponds to "No tone", 2 corresponds to "Simple tone system" and 3 corresponds to "Complex tone system" 

* `$latitude` and `$longitude`: Geographical information of a language 


```{r}
suppressMessages(tone <- read_csv("G:/Shared drives/PSYCH 201a project/Data/tone.csv"))
tone <- as.data.frame(tone)
glimpse(tone)
```
## AUTOTYP

```{r, include = FALSE}
#cleaning
autotyp_1_ <- read_csv("C:/Users/cathe/Downloads/autotyp (1).csv")
autotyp <- autotyp_1_
write.csv(autotyp,'autotyp.csv')
```

### About this dataset

AUTOTYP is a typological database that draws from various existing databases, including WALS. Most of the features in this database are semantic, as opposed to the ones in WALS, which mostly relate to the sound system and grammatical structure of the langauges. AUTOTYP website: http://www.autotyp.uzh.ch/. Documentation: http://www.autotyp.uzh.ch/download/release_2013/autotyp-release_2013.pdf.

The variables in this dataset have the following significance:

* `$LID`: is the language ID code. This is a numeric value. 

* `$ISO639.3`: This is a 3-letter code that refers to the language. These are different from the `$wals_code`.

* `$language`: full language name

* `$alt.language.name`: alternative language name

* `$language.search`: all possible names for the language

* `$lsbranch`: lowest sub-branch (of language family)

* `$ssbranch`: sub-subbranch (of language family)

* `$sbranch`: sub-branch (of language family)

* `$mbranch`: major branch (of language family)

* `$stock`: highest-level language family

* `$alt.stock.name`: all possible names for the language family/group

* `$stock.search`: all possible names for highest-level language family

* `$longitude`: in degrees

* `$latitude`: in degrees

* `$area`: smaller sub-divisions: N Africa, Greater Abyssinia, African Savannah, S Africa, Europe, Greater Mesopotamia, Indic, Inner Asia, Southeast Asia, N Coast Asia, N Coast New Guinea, Interior New Guinea, S New Guinea, Oceania, N Australia, S Australia, Alaska-Oregon, California, Basin and Plains, E North America, Mesoamerica, NE South America, Andean, SE South America

* `$continent`: continent

```{r}
autotyp <- read_csv("autotyp.csv")
autotyp <- autotyp %>% select(-c(X1))
str(autotyp)
```

## Elevation

```{r}
suppressMessages(elev_data_all_languages <- read_csv("G:/Shared drives/PSYCH 201a project/Data/elev_data_all_languages.csv"))
colnames(elev_data_all_languages) <- c("wals.code", "Name", "latitude", "longitude", "country.code", "elev.ft", "elev.m")
elevation <- as.data.frame(elev_data_all_languages)
write.csv(elevation,'elevation.csv')
```

### About this dataset 

This dataset is comprised of the WALS datasets with additional columns for elevation in feet and meters. The elevation data was taken from https://www.gpsvisualizer.com/elevation, which used the Digital Elevation Model Database.



* `$wals.code`: 3-digit code unique to a language (e.g., English = eng) 

* `$Name`: Language name

* `$latitude` and `$longitude`: Geographical information of a language 

* `$elev.ft`: Elevation in feet

* `$elev.m`: Elevation in meters


```{r}
suppressMessages(elevation <- read_csv("G:/Shared drives/PSYCH 201a project/Data/elevation.csv"))
glimpse(elevation)
```

## Humidity

## Temperature

## GDP

```{r, include=FALSE}
#tidying
GDPPC_USD_countries <- read_csv("G:/Shared drives/PSYCH 201a project/Data/GDPPC-USD-countries.csv", col_names = FALSE)
gdp <- GDPPC_USD_countries 
colnames.gdp <- gdp

gdp.2017 <- data.frame(gdp$X1, gdp$X2, gdp$X50)
colnames(gdp.2017) <- c("CountryID", "Country", "GDP.2017")
gdp.2017 <- gdp.2017[-1,]
gdp.2017 <- gdp.2017[-1,]
gdp.2017 <- gdp.2017[-1,]
gdp.2017 <- na.omit(gdp.2017)
write.csv(gdp.2017,'gdp_2017.csv')
```
### About this dataframe

The GDP data came from National Accounts (https://unstats.un.org/unsd/snaama/Downloads). Methodological information:  https://unstats.un.org/unsd/methodology/m49/.

* `$CountryID`: three-digit alphabetical codes assigned by the International Organization for Standardization (ISO). Full list of country codes: https://www.iso.org/obp/ui/#search.

* `$Country`: Country name.

* `$GDP.2017`: GDP in the year 2017 (the most recent year this database provided data for) by country in USD.

```{r}
suppressMessages(gdp_2017 <- read_csv("G:/Shared drives/PSYCH 201a project/Data/gdp_2017.csv"))
glimpse(gdp.2017)
```

## Number of Speakers of Each Language

## Linguistic Diversity by Country

```{r, include=FALSE}
summary_by_country <- read_csv("summary_by_country.csv")
 
View(summary_by_country)
glimpse(summary_by_country)
summary_by_country <- as.data.frame(summary_by_country)
 
#change the div.coverage to a proportion and get rid of percent sign 
summary_by_country$div.coverage = as.numeric(gsub("[\\%,]", "", summary_by_country$div.coverage))
summary_by_country <- summary_by_country %>%
  mutate(div.coverage = div.coverage/100)
summary_by_country$diversity.index <- as.numeric(as.character(summary_by_country$diversity.index))

write.csv(summary_by_country,'summary_by_country_tidied.csv')
```

### About this dataframe

This dataframe comes from Ethnologue (http://www.ethnologue.com/sites/default/files/Ethnologue-21-Global%20Dataset%20Doc.pdf).

* `$lang.count`: number of languages per country
 
* `$lang.percent`: percent of the world's languages spoken in that country

* `$established.langs`: languages more or less "native" to that country

* `$immigrant.langs`: immigrant languages

* `$total.speakers`: population of country

* `$mean.speakers`: mean number of speakers of each language in that country

* `$median.speakers`:  median number of speakers of each language in that country

* `$diversity.index`: range 0-1 how diverse the languages spoken in each country (0 being the lowest, 1 being the highest) (See more info: https://www.nationalgeographic.org/maps/language-diversity-index/)

 
```{r}
suppressMessages(autotyp_geo_info <- read_csv("G:/Shared drives/PSYCH 201a project/Data/autotyp_geo_info.csv")) 
glimpse(autotyp_geo_info)

``` 

# Combining Dataframes

## First combine WALS data ejectives and CV ratio 

```{r}
ejectives_cleaned$eject_d <- ejectives_cleaned$description
ejectives_cleaned$eject_v <- ejectives_cleaned$value
CVratio_cleaned$CV_d <- CVratio_cleaned$description
CVratio_cleaned$CV_v <- CVratio_cleaned$value
wals1 <- merge(ejectives_cleaned, CVratio_cleaned, by = "wals_code", all = TRUE)
str(wals1)

```
Next combine `wals1` with the tone data.

```{r}
tone$tone_d <- tone$description
tone$tone_v <- tone$values
wals2 <- merge(wals1, tone, by.x = "wals_code", by.y = "wals_code", all = TRUE)
str(wals2)
```
`wals_3` now contains the data for the three WALS features. Tidy up the WALS database:

```{r}
wals_3 <- wals2 %>%
  select(c(wals_code, name.x, latitude.x, longitude.x, genus.x, family.x, eject_d, eject_v, CV_d, CV_v, tone_d, tone_v))
wals_3 <- as_data_frame(wals_3)
colnames(wals_3) <- c("wals_code", "name", "latitude", "longitude", "genus", "family", "eject_d", "eject_v", "CV_d", "CV_v", "tone_d", "tone_v")
write.csv(wals_3, "wals_merged.csv")
str(wals_3)
```
These are the relevant feature variables:

* `$eject_d`: This is the description of the ejective value in words. E.g. "no glottalized consonants".

* `$eject_v`: This is the numeric value associated with the ejective feature in each language from WALS. 

* `$CV_d`: This is the description of the consonant-vowel ratio in words. E.g. "High" or "average".

* `$CV_v`: This is the numeric value of the consonant-vowel ratio feature assigned by WALS.

* `$tone_d`: This is the description in words of the tone system in each language. E.g. "simple tone system".

* `$tone_v`: This is the value for the tone system type assigned by WALS.


## Combining WALS with AUTOTYP

Currently, matching these by language types means we lose a few. 

```{r}
suppressMessages(isocodes <- read_csv("isocodes.csv"))
isocodes <- isocodes %>%
  select(c(id, iso_codes, name))
wals_3 <- merge(wals_3, isocodes, by.x = "wals_code", by.y = "id", all = FALSE)
#wals_3[!is.na(wals_3$name.x), ]
# wals_3 <- wals_3 %>% 
#   select(-c("name.y"))

```

```{r}
wals_auto <- merge(wals_3, autotyp, by.x = 'iso_codes', by.y = 'ISO639.3', all = FALSE)
wals_auto <- wals_auto %>% 
  select(-c(latitude.y, longitude.y, language.search, language, alt.language.name, LID, name.y, alt.stock.name, stock.search))
colnames(wals_auto) <- c("iso_codes", "wals_code", "name", "latitude", "longitude", "genus", "family", "eject_d", "eject_v", "CV_d", "CV_v", "tone_d", "tone_v", "lsbranch", "ssbranch", "sbranch", "mbranch", "stock", "area", "continent")
str(wals_auto)
write.csv(wals_auto, "wals_autotyp_merged.csv")
```
These are the current variables:

* `$iso_codes`: ISO code.

* `$wals_code`: WALS code.

* `$name`: language name.

* `$longitude`: in degrees

* `$latitude`: in degrees

* `$genus`: sub-type of language family. 

* `$family`: language family.

* `$eject_d`: This is the description of the ejective value in words. E.g. "no glottalized consonants".

* `$eject_v`: This is the numeric value associated with the ejective feature in each language from WALS. 

* `$CV_d`: This is the description of the consonant-vowel ratio in words. E.g. "High" or "average".

* `$CV_v`: This is the numeric value of the consonant-vowel ratio feature assigned by WALS.

* `$tone_d`: This is the description in words of the tone system in each language. E.g. "simple tone system".

* `$tone_v`: This is the value for the tone system type assigned by WALS.

* `$lsbranch`: lowest sub-branch (of language family)

* `$ssbranch`: sub-subbranch (of language family)

* `$sbranch`: sub-branch (of language family)

* `$mbranch`: major branch (of language family)

* `$stock`: highest-level language family

* `$area`: smaller sub-divisions: N Africa, Greater Abyssinia, African Savannah, S Africa, Europe, Greater Mesopotamia, Indic, Inner Asia, Southeast Asia, N Coast Asia, N Coast New Guinea, Interior New Guinea, S New Guinea, Oceania, N Australia, S Australia, Alaska-Oregon, California, Basin and Plains, E North America, Mesoamerica, NE South America, Andean, SE South America

* `$continent`: continent
