---
title: "4_Analysis"
author: "Catherine Arnett"
date: "11/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

# Replicating Everett's findings

## Replicate Ejective-Elevation Correlation

```{r}
final.data$glot_features_new = final.data$glot_features

final.data$glot_features_new <- ifelse(final.data$glot_features_new == '2 - Ejectives only', 1,
                         ifelse(final.data$glot_features_new == '1 - No glottalized consonants', 0,
                                0))

ejectives.elevation <- lm(data = final.data, glot_features_new ~ elev_m)
summary(ejectives.elevation)
anova(ejectives.elevation)
```


```{r}
final.data$elev_binary <- final.data$elev_m

final.data$elev_binary<- ifelse(final.data$elev_binary >= 1500, 'high',
                         ifelse(final.data$elev_binary < 1500, 'low',
                                'low'))

final.data <- final.data %>%
  mutate(elev_binary = as.character(elev_binary))

ejectives.elevation2 <- lm(data = final.data, glot_features_new ~ elev_binary)
summary(ejectives.elevation2)
```
## Replicate C-V Ratio-Latitude Correlation

```{r}

```

## Replicate Tone-Humidity Correlation

```{r}

```

# Finding alternative correlations

Some factors, which might help explain this apparent correlation between climate and language structure are: a) community size, b) GDP, and c) linguistic diversity of a country or region.


## First look to see whether GDP predicts features

### Ejectives

There is a marginal effect of GDP on whether or not a language 

```{r}
ejectives.GDP <- lm(data = final.data, glot_features_new ~ `2018`)
summary(ejectives.GDP)
```

### CV Ratio


```{r}

```